Switch equivalence and weak prime domains for left-linear DPO rewriting systems

### 2025-04-02 full_version/leftlinear.tex
cose

\begin{remark}\label{rem:seqabs}
Let $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ be an abstraction equivalence between $\der{D}=\{\dder{D}_i\}_{i=0}^1$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^1$. If there is an independence pair between $(i_0, i_1)$ between $\dder{D}_0$ and $\dder{D}_1$, then $(\phi_{D_0}\circ i_1, \phi_{D_1}\circ i_0)$ is an independence pair between $\dder{D}'_0$ and $\dder{D}'_1$. Indeed we have
\[\begin{split}
g'_0\circ \phi_{D_0}\circ i_1&=\phi_{G_1}\circ g_0\circ i_1\\&=\phi_{G_1}\circ m_1\\&= m'_1
\end{split}\qquad \begin{split}
f'_1\circ \phi_{D_1}\circ i_0&=\phi_{G_1}\circ f_1\circ i_0\\&=\phi_{G_1}\circ h_0\\&= h'_1
\end{split}\]
\end{remark}

@@

	A \emph{pre-switch} of $\der{D}$ along $(i_0,i_1)$ is
	a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
	in reverse order, as in the diagram on the right below
	\[
	\xymatrix@R=20pt@C=20pt{
		{L_1} \ar[d]_{m_0'}
		&&  {K_1} \ar@{>->}[ll]_{l_1} \ar[r]^{r_1} \ar[d]_{k_0'}
		&  {R_1} \ar[dr]|(.33)\hole_{h_0'}  \ar@/^.25cm/@{.>}_(.4){i_0'}|(.32)\hole[drrr]
		& & 
		{L_0}\ar[dl]|(.33)\hole^{m_1'} \ar@/_.25cm/@{.>}^(.4){i_1'}[dlll] 
		&  {K_0} \ar@{>->}[l]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
		& & {R_0} \ar[d]^{h_1'} \\		
		{G_0}
		& & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rr]_{g_0'}
		& &  {G'_1} 
		& &  {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rr]_{g_1'}
		& & {G_2}  }
	\]
	such that there is an independence pair $(i_0', i_1')$ between
	$\dder{E}_0$ and $\dder{E}_1$ and 
	\begin{center}   
		$m_0=f_0' \circ i_1'$
		\qquad $h_1=g_1' \circ i_0'$
		\qquad $m_0'= f_0 \circ i_1$
		\qquad $h_1'= g_{1}\circ i_0$.
	\end{center}

@@

A first property of pre-switches is that they are unique up to abstraction equivalence.

@@

@@

\begin{definition}[Switch] 	Let $(\X, \R)$ be a left-linear rewriting system, with $\X$ an $\mathcal{M}$-adhesive category.
	Let also $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation made
	of two sequentially independent derivations $\dder{D}_0$,
	$\dder{D}_1$ with independence pair $(i_0, i_1)$, as in the diagram below, where the cone with vertex $P$ is a pullback and $u_0\colon K_0\to P$ and $u_1\colon K_1\to P$ are the arrows built in \Cref{prop:tec}.
	\[\xymatrix@R=15pt@C=20pt{L_0 \ar[d]_{m_0}&& K_0 \ar @/^.35cm/[ddrr]_(.2){u_0}
		\ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
		\ar@{>}@/^.25cm/[drrr]|(.32)\hole_(.4){i_0}|(.68)\hole
		\ar[dr]|(.33)\hole_{h_0} && L_1 \ar@{>}@/_.25cm/[dlll]^(.4){i_1}|(.68)\hole 
		\ar[dl]|(.33)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar @/_.35cm/[ddll]^(.2){u_1}
		\ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
		D_0 \ar[rr]_{g_0}|(.7)\hole&& G_1 && \ar@{>->}[ll]^{f_1}|(.7)\hole D_1
		\ar[rr]_{g_1}&& G_2 \\ & & & & P \ar[urr]_{p_1} \ar@{>->}[ull]^{p_0}}
	\]

	Let $\der{E}=\{\dder{E}_i\}_{i=0}^1$, depicted below, be a pre-switch for $\der{D}$ along $(i_0,i_1)$.
	\[
	\xymatrix@R=18pt@C=20pt{
		{L_1} \ar[d]_{m_0'}
		&&  {K_1} \ar@{>->}[ll]_{l_1} \ar[r]^{r_1} \ar[d]_{k_0'}
		&  {R_1} \ar[dr]|(.33)\hole_{h_0'}  \ar@/^.25cm/@{>}_(.4){i_0'}|(.32)\hole[drrr]
		& & 
		{L_0}\ar[dl]|(.33)\hole^{m_1'} \ar@/_.25cm/@{>}^(.4){i_1'}[dlll] 
		&  {K_0} \ar@{>->}[l]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
		& & {R_0} \ar[d]^{h_1'} \\		
		{G_0}
		& & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rr]_{g_0'}
		& &  {G'_1} 
		& &  {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rr]_{g_1'}
		& & {G_2}  }
	\]

@@

\begin{proposition}\label{prop:pretosw} 	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^1$ be a derivation and suppose
	that both $\der{E}=\{\dder{E}_i\}_{i=0}^1$ and
	$\der{F}=\{\dder{F}_i\}_{i=0}^1$ are pre-switches of $\dder{D}_0$ and
	$\dder{D}_1$ along the same independence pair $(i_0, i_1)$. Let $\{\phi_X\}_{X\in \Deltamin(\der{E})}$ be the abstraction equivalence built in \Cref{thm:switch_uni}. If $\der{F}$ is a switch then also $\der{E}*\phi_{G_2}$ is one.
\end{proposition}
\begin{proof}
	contenuto...
\end{proof}

@@

		Moreover
		\[\begin{split}
			f'_0\circ z_0&=f'_0\circ t_0\circ v\\&=f'_0\circ \phi_{Q_0}\circ q_0\\&=a_0\circ q_0\\&=f_0\circ p_0
		\end{split} \qquad \begin{split}
		g'_1\circ z_1=g'_1\circ t_1\circ v\\&=g'_1\circ \phi_{Q_1}\circ q_1\\&=\phi_{G_2}\circ b_1 \circ q_1\\&=\phi_{G_2}\circ g_1\circ p_1\\&=f_0\circ p_0
		\end{split} \]

>>>>

\begin{remark}\label{rem:seqabs}
Let $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ be an abstraction equivalence between $\der{D}=\{\dder{D}_i\}_{i=0}^1$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^1$. If there is an independence pair between $(i_0, i_1)$ between $\dder{D}_0$ and $\dder{D}_1$, then $(\phi_{D_1}\circ i_0, \phi_{D_0}\circ i_1)$ is an independence pair between $\dder{D}'_0$ and $\dder{D}'_1$. Indeed we have
\[\begin{split}
g'_0\circ \phi_{D_0}\circ i_1&=\phi_{G_1}\circ g_0\circ i_1\\&=\phi_{G_1}\circ m_1\\&= m'_1
\end{split}\qquad \begin{split}
f'_1\circ \phi_{D_1}\circ i_0&=\phi_{G_1}\circ f_1\circ i_0\\&=\phi_{G_1}\circ h_0\\&= h'_1
\end{split}\]
\end{remark}

@@

	A \emph{pre-switch} of $\der{D}$ along $(i_0,i_1)$ is
	a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
	in reverse order, as in the diagram on the right below
	\[
	\xymatrix@R=20pt@C=20pt{
		{L_1} \ar[d]_{m_0'}
		&&  {K_1} \ar@{>->}[ll]_{l_1} \ar[r]^{r_1} \ar[d]_{k_0'}
		&  {R_1} \ar[dr]|(.33)\hole_{h_0'}  \ar@/^.25cm/@{.>}_(.4){i_0'}|(.32)\hole[drrr]
		& & 
		{L_0}\ar[dl]|(.33)\hole^{m_1'} \ar@/_.25cm/@{.>}^(.4){i_1'}[dlll] 
		&  {K_0} \ar@{>->}[l]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
		& & {R_0} \ar[d]^{h_1'} \\		
		{G_0}
		& & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rr]_{g_0'}
		& &  {G'_1} 
		& &  {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rr]_{g_1'}
		& & {G_2}  }
	\]
	such that there is an independence pair $(i_0', i_1')$ between
	$\dder{E}_0$ and $\dder{E}_1$ and 
\[
	m_0=f_0' \circ i_1' \qquad 
	h_1=g_1' \circ i_0' 
		\qquad m_0'= f_0 \circ i_1
		\qquad h_1'= g_{1}\circ i_0\]

@@

We start by examining how pre-switches interact with abstraction equivalence. A first property that we can establish in that direction is uniqueness of pre-switches up to abstraction equivalence.

@@

Another useful facts that we can establish is that being a pre-switch is invariant under abstraction equivalence, as made precise by the following lemma.

\begin{lemma}\label{lem:abs}
	Let $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be the following derivation:

	\[\xymatrix@R=20pt@C=20pt{L_0 \ar[d]_{m_0}&& K_0
		\ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
		\ar@{>}@/^.25cm/[drrr]|(.32)\hole_(.4){i_0}
		\ar[dr]|(.33)\hole_{h_0} && L_1 \ar@{>}@/_.25cm/[dlll]^(.4){i_1}
		\ar[dl]|(.33)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
		\ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
		D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
		\ar[rr]_{g_1}&& G_2}
	\]
	and suppose that the derivation $\der{E}$ below is a pre-switch for $\der{D}$ along $(i_0,i_1)$.
	\[
	\xymatrix@R=20pt@C=20pt{
		{L_1} \ar[d]_{m_0'}
		&&  {K_1} \ar@{>->}[ll]_{l_1} \ar[r]^{r_1} \ar[d]_{k_0'}
		&  {R_1} \ar[dr]|(.33)\hole_{h_0'}  \ar@/^.25cm/@{>}_(.4){i_0'}|(.32)\hole[drrr]
		& & 
		{L_0}\ar[dl]|(.33)\hole^{m_1'} \ar@/_.25cm/@{>}^(.4){i_1'}[dlll] 
		&  {K_0} \ar@{>->}[l]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
		& & {R_0} \ar[d]^{h_1'} \\		
		{G_0}
		& & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rr]_{g_0'}
		& &  {G'_1} 
		& &  {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rr]_{g_1'}
		& & {G_2}  }
	\]

	Then the following hold true:
	\begin{enumerate}
		\item if $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ is an abstraction equivalence between $\der{D}$ and another derivation $\der{F}=\{\dder{F}_i\}_{i=0}^1$, then $\phi^{-1}_{G_0}*\der{E}*\phi_{G_2}$ is a pre-switch for $\der{F}$ along $(\phi_{D_1}\circ i_0, \phi_{D_0}\circ i_1 )$;
		\item if $\{\psi_X\}_{X\in \Deltamin(\der{E})}$ is an abstraction equivalence between $\der{E}$ and $\der{G}=\{\dder{G}_{i}\}_{i=0}^1$, then $\psi_{G_0}*\der{G}*\psi^{-1}_{G_2}$ is a pre-switch for $\der{D}$ along $(i_0, i_1)$.
	\end{enumerate}
	\begin{remark}Recall that, by \Cref{rem:seqabs}, $(\phi_{D_1}\circ i_0, \phi_{D_0}\circ i_1 )$ is an independence pair between $\dder{F}_0$ and $\dder{F}_1$.
	\end{remark}
\end{lemma}
\begin{proof}\begin{enumerate}
		\item To prove this, it is enough to compute:
		\[\begin{split}
			m_{\der{F},0}=&\phi_{G_0}\circ m_0\\=&\phi_{G_0}\circ f'_0\circ i'_1\\\phi_{G_0}\circ	m'_0=&\phi_{G_0}\circ f_0\circ i_1\\=&f_{\der{F},0}\circ \phi_{D_{0}}\circ i_1
		\end{split} \quad \begin{split}
	h_{\der{F},1}=&\phi_{G_2}\circ h_1\\=&\phi_{G_2}\circ g_1' \circ i_0' \\	\phi_{G_2}\circ h'_1=&\phi_{G_2}\circ g_1\circ i_0\\=&g_{\der{F}, 1}\circ \phi_{D_1}\circ i_0\end{split} 
	\]

		\item

		And we can conclude. \qedhere 
	\end{enumerate}
\end{proof}

@@

\begin{definition}[Switch] 	Let $(\X, \R)$ be a left-linear rewriting system, with $\X$ an $\mathcal{M}$-adhesive category.
	Let also $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation made
	of two sequentially independent derivations $\dder{D}_0$,
	$\dder{D}_1$ with independence pair $(i_0, i_1)$, as in the diagram below, where the cone with vertex $P$ is a pullback and $u_0\colon K_0\to P$ and $u_1\colon K_1\to P$ are the arrows built in \Cref{prop:tec}.
	\[\xymatrix@R=20pt@C=20pt{L_0 \ar[d]_{m_0}&& K_0 \ar @/^.35cm/[ddrr]_(.2){u_0}
		\ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
		\ar@{>}@/^.25cm/[drrr]|(.32)\hole_(.4){i_0}|(.68)\hole
		\ar[dr]|(.33)\hole_{h_0} && L_1 \ar@{>}@/_.25cm/[dlll]^(.4){i_1}|(.68)\hole 
		\ar[dl]|(.33)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar @/_.35cm/[ddll]^(.2){u_1}
		\ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
		D_0 \ar[rr]_{g_0}|(.7)\hole&& G_1 && \ar@{>->}[ll]^{f_1}|(.7)\hole D_1
		\ar[rr]_{g_1}&& G_2 \\ & & & & P \ar[urr]_{p_1} \ar@{>->}[ull]^{p_0}}
	\]

	Let $\der{E}=\{\dder{E}_i\}_{i=0}^1$, depicted below, be a pre-switch for $\der{D}$ along $(i_0,i_1)$.
	\[
	\xymatrix@R=20pt@C=20pt{
		{L_1} \ar[d]_{m_0'}
		&&  {K_1} \ar@{>->}[ll]_{l_1} \ar[r]^{r_1} \ar[d]_{k_0'}
		&  {R_1} \ar[dr]|(.33)\hole_{h_0'}  \ar@/^.25cm/@{>}_(.4){i_0'}|(.32)\hole[drrr]
		& & 
		{L_0}\ar[dl]|(.33)\hole^{m_1'} \ar@/_.25cm/@{>}^(.4){i_1'}[dlll] 
		&  {K_0} \ar@{>->}[l]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
		& & {R_0} \ar[d]^{h_1'} \\		
		{G_0}
		& & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rr]_{g_0'}
		& &  {G'_1} 
		& &  {D_1'} \ar@{>->}[ll]^{f_1'} \ar[rr]_{g_1'}
		& & {G_2}  }
	\]

@@

\begin{proposition}\label{prop:pretosw} 	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^1$ be a derivation and suppose
	that both $\der{E}=\{\dder{E}_i\}_{i=0}^1$ and
	$\der{F}=\{\dder{F}_i\}_{i=0}^1$ are pre-switches of $\dder{D}_0$ and
	$\dder{D}_1$ along the same independence pair $(i_0, i_1)$. Let $\{\phi_X\}_{X\in \Deltamin(\der{E})}$ be the abstraction equivalence built in \Cref{thm:switch_uni}. If $\der{F}$ is a switch then also $\der{E}*\phi_{G_2}$ is one.
\end{proposition}
\begin{proof}
Suppose that $\der{D}$ is given by the following derivation.

	By \Cref{thm:switch_uni} we have a diagram as the one below.
	\[\xymatrix@C=22pt{G_{0} && \ar@{>->}[ll]_{f_{\der{F},0}} D_{\der{F},0} \ar[rr]^{g_{\der{F},0}}&& G_{\der{F},1} && \ar@{>->}[ll]_{f_{\der{F},1}} D_{\der{F},1} \ar[rr]^{g_{\der{F},1}}&& G_{2}\\L_1 \ar[d]^{f_0\circ i_1} \ar[u]_{f_0\circ i_1}&& K_{1} \ar[d]^{k_{\der{E},0}} \ar[u]_{k_{\der{F},0}}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_{1} \ar@/^.35cm/[drrr]|(.315)\hole_(.4){j_0} \ar[dr]|(.3)\hole_{h_{\der{E},0}} \ar@/_.35cm/[urrr]|(.315)\hole^(.4){a_0} \ar[ur]|(.3)\hole^{h_{\der{F},0}}& &L_{0} \ar@/_.35cm/[dlll]^(.4){j_1} \ar[dl]|(.3)\hole^{m_{\der{E},1}} \ar@/^.35cm/[ulll]_(.4){a_1} \ar@/_.3cm/[ul]|(.252)\hole_(.4){m_{\der{F},1}}& K_{0} \ar[d]_{k_{\der{E},1}} \ar[u]^{k_{\der{F},1}}\ar@{>->}[l]_{l_{0}} \ar[rr]^{r_0} && R_{0} \ar[d]_{g_1\circ i_0} \ar[u]^{g_1\circ i_0}\\ \ar@/^.5cm/[uu]^{\id{G_0}} G_{0} && \ar[ll]^{f_{\der{E},0}} D_{\der{E},0} \ar@{>}@/^.5cm/[uu]^(.33){\phi_{D_{\der{E},0}}}|\hole \ar[rr]_{g_{\der{E},0}}&& G_{\der{E},1} \ar@{>}[uu]_(.76){\phi_{G_{\der{E}, 1}}}|(.45)\hole |(.55)\hole&& \ar@{>->}[ll]^{f_{\der{E},1}} D_{\der{E},1} \ar@{>}@/_.5cm/[uu]_(.33){\phi_{D_{\der{E},1}}}|\hole\ar[rr]_{g_{\der{E},1}}&& G_{2}\ar@{>}@/_.5cm/[uu]_{\phi_{G_2}}}\]

	Moreover, by definition $\der{E}*\phi_{G_2}$

@@

		Moreover
		\[\begin{split}
			f'_0\circ z_0&=f'_0\circ t_0\circ v\\&=f'_0\circ \phi_{Q_0}\circ q_0\\&=a_0\circ q_0\\&=f_0\circ p_0
		\end{split} \qquad \begin{split}
		g'_1\circ z_1&=g'_1\circ t_1\circ v\\&=g'_1\circ \phi_{Q_1}\circ q_1\\&=\phi_{G_2}\circ b_1 \circ q_1\\&=\phi_{G_2}\circ g_1\circ p_1\\&=f_0\circ p_0
		\end{split} \]
